---
title: "data_exploration_initial"
output: html_document
---

Set the filename for your local file location

```{r}
fileName <-"/home/dev/CSCD429_Project/DJIA_table.csv"
```

Read in the stock data and transform the Date column into actual dates
```{r}
library(dplyr)
stocks <- read.csv(file=fileName, header=TRUE, sep=",", stringsAsFactors = FALSE)
stocks <- mutate(stocks, Date=as.Date(Date))
```

Set the cut off date we want to use to separate our training and testing data
```{r}
cutOffDate <- as.Date("2016-01-01")
```
Separate the data
```{r}
stock_train <- subset(stocks,Date<cutOffDate)
stock_test <- subset(stocks, Date>=cutOffDate)
```

Export the training and testing data so we can share it - currently commented out since we don't need to do this every time
```{r}
#write.csv(stock_train, file="/home/dev/CSCD429_Project/DJIA_training.csv", row.names=FALSE, na="", sep=",")
#write.csv(stock_test, file="/home/dev/CSCD429_Project/DJIA_test.csv", row.names=FALSE, na="", sep=",")
```

Addings some new columns: the daily spread of the data (high - low), the movement within the day itself (open - close), the difference between the previous close and the current close, and the difference bewteen the previous close and the current opening
```{r}
stock_train$Spread <- stock_train$High - stock_train$Low
stock_train$DailyMove <- stock_train$Open - stock_train$Close

dailyDiffs <- function(x) {
  for (i in 1:(dim(x)[1] - 1)) {
    x$CloseDiff[i] <- x$Close[i+1] - x$Close[i]
    x$ONDiff[i] <- x$Close[i+1] - x$Open[i]
    x$SpreadPercent[i] <- x$Spread[i] / x$Close[i+1]
    x$DailyMovePercent[i] <- x$DailyMove[i] / x$Close[i+1]
    x$CloseDiffPercent[i] <- x$CloseDiffPercent[i] / x$Close[i+1]
    x$ONDiffPercent[i] <- x$ONDiff[i] / x$Close[i+1]
  }
  return(x)
}
stock_train <- dailyDiffs(stock_train)
```

Now lets add some additional data, standard deviation, mean, median
```{r}
stock_stats <- data.frame(colnames(stock_train)[-1])
colnames(stock_stats)[1] <- 'Name'
rownames(stock_stats) <- stock_stats$Name

std <- c();
train_mean <- c()
train_med <- c() 
for (i in 1:(dim(stock_stats)[1])) {
  std[i] <- (var(stock_train[[as.character(stock_stats$Name[i])]], na.rm=TRUE)^(1/2))
  train_mean[i] <- (mean(stock_train[[as.character(stock_stats$Name[i])]], na.rm=TRUE))
  train_med[i] <- (median(stock_train[[as.character(stock_stats$Name[i])]], na.rm=TRUE))
}

stock_stats$StandardDev <- std
stock_stats$Mean <- train_mean
stock_stats$Median <- train_med
```

Now lets do some labeling of the movements and values. Right now these cuts are arbitrary, we should decice where and how we want to categorize
```{r}
rownames(stock_stats) <- stock_stats$Name

getLabel <- function(value, standardDev) {
  label <- ""
  if (is.na(value) || is.na(standardDev)) {
    label <- NULL
  } else if (value < -3 * standardDev) {
    label <- "Down Extreme"
  } else if (value < -1.5 * standardDev) {
    label <- "Down Large"
  } else if (value < -0.1 * standardDev) {
    label <- "Down"
  } else if (value < 0.1 * standardDev) {
    label <- "No Move"
  } else if (value < 1.5 * standardDev) {
    label <- "Up"
  } else if (value < 3 * standardDev) {
    label <- "Up Large"
  } else {
    label <- "Up Extreme"
  }
  
  return(label)
}

for (i in 1:(dim(stock_train)[1] - 1)) {
  stock_train$DailyMoveCat[i] <- getLabel(stock_train$DailyMove[i], stock_stats["DailyMove","StandardDev"])
  stock_train$CloseDiffCat[i] <- getLabel(stock_train$CloseDiff[i], stock_stats["CloseDiff","StandardDev"])
  stock_train$ONDiffCat[i] <- getLabel(stock_train$ONDiff[i], stock_stats["ONDiff","StandardDev"])
}

labels <- c("Down Extreme", "Down Large", "Down", "No Move", "Up", "Up Large", "Up Extreme")
stock_train$DailyMoveCat <-factor(stock_train$DailyMoveCat, levels=labels, ordered = TRUE)
stock_train$CloseDiffCat <-factor(stock_train$CloseDiffCat, levels=labels, ordered = TRUE)
stock_train$ONDiffCat <-factor(stock_train$ONDiffCat, levels=labels, ordered = TRUE)

table(stock_train$DailyMoveCat)
table(stock_train$CloseDiffCat)
table(stock_train$ONDiffCat)
```

Finally here are just some histograms to visualize the data

```{r}
hist(stock_train$Spread, breaks=50)
hist(stock_train$DailyMove, breaks=50)
hist(stock_train$CloseDiff, breaks=50)
hist(stock_train$ONDiff, breaks=50)
```
